---
title: "Data Science for Business - Time Series Forcasting Part 1: EDA & Data Preparation"
author: "Dr. Shirin Glander"
date: "May 22, 2017"
output: html_document
---

http://archive.ics.uci.edu/ml/datasets/Online+Retail

This is a transnational data set which contains all the transactions occurring between 01/12/2010 and 09/12/2011 for a UK-based and registered non-store online retail.The company mainly sells unique all-occasion gifts. Many customers of the company are wholesalers.

Attribute Information:

InvoiceNo: Invoice number. Nominal, a 6-digit integral number uniquely assigned to each transaction. If this code starts with letter 'c', it indicates a cancellation. 
StockCode: Product (item) code. Nominal, a 5-digit integral number uniquely assigned to each distinct product. 
Description: Product (item) name. Nominal. 
Quantity: The quantities of each product (item) per transaction. Numeric.	
InvoiceDate: Invice Date and time. Numeric, the day and time when each transaction was generated. 
UnitPrice: Unit price. Numeric, Product price per unit in sterling. 
CustomerID: Customer number. Nominal, a 5-digit integral number uniquely assigned to each customer. 
Country: Country name. Nominal, the name of the country where each customer resides.


Relevant Papers:

The evolution of direct, data and digital marketing, Richard Webber, Journal of Direct, Data and Digital Marketing Practice (2013) 14, 291–309. 
Clustering Experiments on Big Transaction Data for Market Segmentation, 
Ashishkumar Singh, Grace Rumantir, Annie South, Blair Bethwaite, Proceedings of the 2014 International Conference on Big Data Science and Computing. 
A decision-making framework for precision marketing, Zhen You, Yain-Whar Si, Defu Zhang, XiangXiang Zeng, Stephen C.H. Leung c, Tao Li, Expert Systems with Applications, 42 (2015) 3357–3367.



Citation Request:

Daqing Chen, Sai Liang Sain, and Kun Guo, Data mining for the online retail industry: A case study of RFM model-based customer segmentation using data mining, Journal of Database Marketing and Customer Strategy Management, Vol. 19, No. 3, pp. 197–208, 2012 (Published online before print: 27 August 2012. doi: 10.1057/dbm.2012.17).


http://r4ds.had.co.nz/

R for Data Science
Garrett Grolemund
Hadley Wickham


```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidyquant)
library(modelr)
library(gridExtra)
library(grid)
```

```{r eval=FALSE}
retail <- read_csv("OnlineRetail.csv",
                   col_types = cols(
  InvoiceNo = col_character(),
  StockCode = col_character(),
  Description = col_character(),
  Quantity = col_integer(),
  InvoiceDate = col_datetime("%m/%d/%Y %H:%M"),
  UnitPrice = col_double(),
  CustomerID = col_integer(),
  Country = col_character()
)) %>%
  mutate(day = parse_date(format(InvoiceDate, "%Y-%m-%d")),
         day_of_week = wday(day, label = TRUE),
         time = parse_time(format(InvoiceDate, "%H:%M")),
         month = format(InvoiceDate, "%m"),
         income = Quantity * UnitPrice,
         income_return = ifelse(Quantity > 0, "income", "return"))
```

```{r echo=FALSE, eval=FALSE}
save(retail, file = "retail.RData")
```

```{r echo=FALSE}
load("retail.RData")
```

<br>

## Exploratory Data Analysis (EDA)

```{r fig.width=10, fig.height=4}
p1 <- retail %>%
  filter(Country == "United Kingdom") %>%
  ggplot(aes(x = Country, fill = income_return)) +
    geom_bar(alpha = 0.8) +
    scale_fill_manual(values = palette_light()) +
    theme_tq() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    guides(fill = FALSE) +
    labs(x = "")

p2 <- retail %>%
  filter(Country != "United Kingdom") %>%
  ggplot(aes(x = Country, fill = income_return)) +
    geom_bar(alpha = 0.8) +
    scale_fill_manual(values = palette_light()) +
    theme_tq() +
    theme(legend.position = "right") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(x = "",
         fill = "")

grid.arrange(p1, p2, widths = c(0.2, 0.8))
```

```{r fig.width=8, fig.height=4}
retail %>%
  ggplot(aes(x = day, color = income_return)) +
    facet_grid(income_return ~ ., scales = "free") +
    geom_freqpoly(bins = 100, size = 1, alpha = 0.8) +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    guides(color = FALSE) +
    labs(title = "Number of purchases/returns per day",
         x = "")
```

```{r fig.width=8, fig.height=3}
retail %>%
  ggplot(aes(x = day, y = ..density.., color = income_return)) +
    geom_freqpoly(size = 1, alpha = 0.8, bins = 100) +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    labs(title = "Density of purchases/returns per day",
         x = "",
         color = "")
```

```{r fig.width=8, fig.height=4}
retail %>%
  group_by(day, income_return) %>%
  summarise(sum_income = sum(income)) %>%
  ggplot(aes(x = day, y = sum_income, color = income_return)) +
    facet_grid(income_return ~ ., scales = "free") +
    geom_ref_line(h = 0, colour = "grey") +
    geom_line(size = 1, alpha = 0.8) +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    guides(color = FALSE) +
    labs(title = "Income from purchases/returns per day",
         x = "",
         y = "sum of income/losses",
         color = "")
```

```{r fig.width=8, fig.height=4}
retail %>%
  group_by(time, income_return) %>%
  summarise(sum_income = sum(income)) %>%
  ggplot(aes(x = time, y = sum_income, color = income_return)) +
    facet_grid(income_return ~ ., scales = "free") +
    geom_ref_line(h = 0, colour = "grey") +
    geom_line(size = 1, alpha = 0.8) +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    guides(color = FALSE) +
    labs(title = "Income from purchases/returns per time of day",
         x = "time of day",
         y = "sum of income/losses",
         color = "")
```

```{r fig.width=8, fig.height=5}
retail %>%
  ggplot(aes(x = time, y = day)) +
    stat_bin2d(alpha = 0.8, bins = 25, color = "white") +
    scale_fill_gradientn(colours = c(palette_light()[[1]], palette_light()[[2]])) +
    theme_tq() +
    theme(legend.position = "right") +
    labs(title = "Purchases/returns per day and time")
```

```{r fig.width=8, fig.height=5}
retail %>%
  mutate(day2 = format(InvoiceDate, "%d")) %>%
  group_by(month, day2) %>%
  summarise(sum_income = sum(income)) %>%
  ggplot(aes(x = month, y = day2, fill = sum_income)) +
    geom_tile(alpha = 0.8, color = "white") +
    scale_fill_gradientn(colours = c(palette_light()[[1]], palette_light()[[2]])) +
    theme_tq() +
    theme(legend.position = "right") +
    labs(title = "Net income per month and day",
         y = "day of the month",
         fill = "net sum of income")
```

<br>

## Preparing data for modeling

- Which customers are repeat customers?

n: number of different items purchased/returned per customer per day.

repeat_customer: TRUE if the customer ID has been recorded on more than one day (i.e., they have made multiple transactions during the year of recording)

Apparently, none... at least if customer id is given per customer, not transaction.

```{r}
rep_customer <- retail %>%
  group_by(day, CustomerID) %>%
  summarise(n = n()) %>%
  mutate(repeat_customer = ifelse(duplicated(CustomerID), TRUE, FALSE))

length(which(rep_customer$repeat_customer == TRUE))
```

<br>

- How many purchases/returns per customer and day?

n: number of different items purchased/returned per customer per day

sum_it: sum of items (quantity) purchased/returned per customer per day

sum_in: mean income/loss per customer per day

mean_in_cust: mean income from all customers per day

```{r}
customer_purch <- retail %>%
  group_by(day, CustomerID, income_return) %>%
  summarise(n = n(),
            sum_it = sum(Quantity),
            sum_in = sum(income)) %>%
  group_by(day) %>%
  summarise(mean_in_cust = mean(sum_in))
```

<br>

- How many items are purchases/returned per day?

```{r}
income_return <- retail %>%
  group_by(day, income_return) %>%
  summarise(sum = sum(Quantity)) %>%
  spread(key = income_return, value = sum)
```

<br>

- How many items are purchased/returned per country?

```{r}
country_purch <- retail %>%
  mutate(Country2 = ifelse(Country == "United Kingdom", "uk", "other_country")) %>%
  group_by(day, Country2) %>%
  summarise(sum = sum(Quantity)) %>%
  spread(key = Country2, value = sum)
```

<br>

- Net income summaries:

```{r}
income <- retail %>%
  group_by(day) %>%
  summarise(sum_income = sum(income),
            mean_income = mean(income),
            sum_quantity = sum(Quantity),
            mean_quantity = mean(Quantity))
```

<br>

- Income from purchases:

```{r}
purchases <- retail %>%
  filter(income > 0) %>%
  group_by(day) %>%
  summarise(sum_income_purch = sum(income),
            mean_income_purch = mean(income),
            sum_quantity_purch = sum(Quantity),
            mean_quantity_purch = mean(Quantity))
```

<br>

- Loss from returns:

```{r}
returns <- retail %>%
  filter(income < 0) %>%
  group_by(day) %>%
  summarise(sum_income_return = sum(income),
            mean_income_return = mean(income),
            sum_quantity_return = sum(Quantity),
            mean_quantity_return = mean(Quantity))
```

<br>

- Mean price per units sold per day

This is a bit more complicated than thought because some items are sold multiple times per time and unit prices of the same item change between transactions and days.

```{r}
temp <- distinct(select(retail, day, StockCode, UnitPrice)) %>%
  mutate(temp = paste(day, StockCode, sep = "_")) %>%
  select(temp, UnitPrice)

mean_unit_price <- retail %>%
  filter(income_return == "income") %>%
  group_by(day, StockCode) %>%
  summarise(n = n()) %>%
  mutate(temp = paste(day, StockCode, sep = "_")) %>%
  left_join(temp, by = "temp") %>%
  group_by(day, StockCode) %>%
  summarise(mean = mean(UnitPrice)) %>%
  group_by(day) %>%
  summarise(mean_unit_price = mean(mean))
```

<br>

- Combining data

```{r eval=FALSE}
retail_p_day <- distinct(select(retail, day, day_of_week, month)) %>%
  left_join(income, by = "day") %>%
  left_join(mean_unit_price, by = "day") %>%
  left_join(purchases, by = "day") %>%
  left_join(returns, by = "day") %>%
  left_join(customer_purch, by = "day") %>%
  left_join(income_return, by = "day") %>%
  left_join(country_purch, by = "day") %>%
  mutate(diff_sum_income = sum_income - lag(sum_income),
         season = ifelse(month %in% c("03", "04", "05"), "spring",
                         ifelse(month %in% c("06", "07", "08"), "summer",
                                ifelse(month %in% c("09", "10", "11"), "fall", "winter"))))
```

```{r echo=FALSE, eval=FALSE}
save(retail_p_day, file = "retail_p_day.RData")
```

```{r echo=FALSE}
load("retail_p_day.RData")
```

```{r}
meta <- tibble(colnames_retail_p_day = colnames(retail_p_day)) %>%
  mutate(description = c(
    "day in YYYY-MM-DD",
    "weekday (Sun - Fri, there are no Sat in the dataset)",
    "month (as month number)",
    
    "sum of net income per day (all purchases & losses per day combined)",
    "mean net income per day (all purchases & losses per day combined)",
    "sum of net quantities sold/returned per day (all purchases & returns per day combined)",
    "mean net quantities sold/returned per day (all purchases & returns per day combined)",
    
    "mean price per unit sold (returns excluded)",
    
    "sum of income from purchases per day (losses excluded)",
    "mean income from purchases per day (losses excluded)",
    "sum of quantities from purchases per day (losses excluded)",
    "mean quantities from purchases per day (losses excluded)",
    
    "sum of losses from returns per day (purchases excluded)",
    "mean losses from returns per day (purchases excluded)",
    "sum of quantities from returns per day (purchases excluded)",
    "mean quantities from returns per day (purchases excluded)",
    
    "mean net income from all customers per day",
    
    "sum of items (quantities) purchased per day (returns excluded)",
    "sum of items (quantities) returned per day (purchases excluded)",
    
    "net sum of items (quantities) purchased per day from countries other than the UK",
    "net sum of items (quantities) purchased per day from the UK",
    
    "difference in sum of net income (purchases - returns) to previous day",
    "season"
    ))

meta
```

```{r fig.width=15, fig.height=30, warning=FALSE, message=FALSE}
retail_p_day %>%
  # remove last value because it is so extreme
  filter(day != max(retail_p_day$day)) %>%
  gather(x, y, sum_income:diff_sum_income) %>%
  ggplot(aes(x = day, y = y)) +
    facet_wrap(~ x, scales = "free", ncol = 2) +
    geom_line(alpha = 0.8, color = palette_light()[[1]]) +
    geom_point(aes(color = day_of_week)) +
    geom_smooth() +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    labs(x = "",
         y = "",
         color = "day of the week")
```

------------------

<br>

```{r }
sessionInfo()
```
